---
title: "CPSC444_Project"
author: "Yuting Qiu"
date: "11/2/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F)
```


You are expected to submit a word or PDF file (no more than 2 pages). 
Name your file as YourNetID_CPSC444_ProjectProposal
The proposal is consisting of 4 parts:
Introduction:
The introduction part should introduce the background and justify your research  and include at least one well-defined research goal.
These goals can consist of:
Generating a dataset
Model the effect of explanatory variables or treatments
Spatial predictions
Datasets:
The data part should include a brief description of your study area, the source of your data, and explanation of the variables you are going to use. 
Methods:
The methods part should explain clearly the analysis you are planning to apply. Make sure the  methods are feasible. Describe the expected result from the analysis. 
References:
Use proper citations for the background, methods and if needed from the dataset.


```{r}
library(readxl)
library(nlme)
library(ggplot2)
library(dplyr)
library(tidyr)
library(sp)
library(lme4)
library(gstat)
library(gridExtra)
```

```{r}
bi_18 <- read_excel("bi.18.wide.xlsx")
bs_18 <- read_excel("bs.18.wide.xlsx")

bi_18 %>%
  select(plot_id,plot,row_num,range_num,rep,audpc) %>% 
  mutate(rep = factor(rep), plot = factor(plot), audpc=as.numeric(audpc)) -> bi_18.1
bs_18 %>%
  select(plot_id,plot,row_num,range_num,rep,audpc) %>% 
  mutate(rep = factor(rep), plot = factor(plot), audpc=as.numeric(audpc)) -> bs_18.1


bi_18.1 <- na.omit(bi_18.1)
bs_18.1 <- na.omit(bs_18.1)
```


plot the arrangement of the field
```{r}
ggplot(bi_18.1, aes(x = range_num, y = row_num)) +
  geom_tile(aes(fill = plot)) # to create a heatmap
  theme_bw() + theme(panel.grid = element_blank())
```


```{r}
m0_bi_PGR <- lme(audpc ~ plot, bi_18.1, random = ~ 1 | rep, na.action = "na.omit")
anova(m0_bi_PGR, type = "marginal")
plot(m0_bi_PGR)
# there is trend left in the residuals
```


```{r}
# Getting residuals from the model
bi_18.1$resid_m0 <- residuals(m0_bi_PGR, type = "pearson")
bi_18.1_sp <- bi_18.1
coordinates(bi_18.1_sp) <- ~ range_num + row_num # give it coordiantes

# Spatial distribution
p_resid_m0 <- ggplot(bi_18.1, aes(x = range_num, y = row_num)) +
  geom_tile(aes(fill = resid_m0)) +
  theme_bw() + theme(panel.grid = element_blank()) +
  scale_fill_gradient(low = "darkblue", high = "orange") +
  ggtitle("Residual plot of 2018_BI population") + 
  labs(fill = "Residuals of model1") +
  xlab("row number") +
  ylab("range number")
  
p_resid_m0

# pretty clear spatial trend in this plot
# there are more negative values at the small row number
```


```{r}
# Getting residuals from the model: no trend
vg0_bi<- Variogram(m0_bi_PGR, form = ~ range_num + row_num, resType = "pearson", maxDist = 50)
plot(vg0_bi, smooth = F, main = "BI population variogram")
```

This variogram shows some structure, however it was not computed from a model including Xft and Yft as fixed predictors, thus we can not tell how much of that spatial structure can be attributed to the trend or it is structured random variation. To find this out we have to look at the variogram from a model including Xft and Yft as predictors.


```{r}
# Split-plot model with range and row as fixed predictors (for detrending)
m1_bi_PGR <- lme(audpc ~ plot, bi_18.1, random = ~ 1|rep + 1|range_num + 1|row_num, na.action = "na.omit")
m1_bi_PG_range <- lme(audpc ~ plot, bi_18.1, random = ~ 1|rep + 1|range_num, na.action = "na.omit")
anova(m1_bi_PG_range, m1_bi_PGR)

# Getting residuas from model 2
bi_18.1$resid_m1 <- residuals(m1_bi_PGR, type = "pearson")

# Getting variogram from the model
vg1_bi <- Variogram(m1_bi_PGR, form = ~ range_num + row_num, resType = "pearson", maxDist = 50)
plot(vg1_bi, smooth= F)
# the variogram seems to be much better
```

Now we can compare the effect of adding range and row for detrending purposes.
```{r}
p_resid_m1 <- ggplot(bi_18.1, aes(x = range_num, y = row_num)) +
  geom_tile(aes(fill = resid_m1)) +
  scale_fill_gradient(low = "darkblue", high = "orange") + 
  theme_bw() + theme(panel.grid = element_blank())
p_resid_m1

grid.arrange( 
  p_resid_m0 + labs(title = "with trend"),
  p_resid_m1 + labs(title = "without trend"),
  nrow = 1,
  top="BI residual plots"
)

# Comparing variograms
grid.arrange(
  plot(vg0_bi, smooth = F, main = "with trend"),
  plot(vg1_bi, smooth  =F, main = "without trend"),
  nrow = 1,
  top = "BI semivariogram plots"
)
```



```{r}
# Defining intial parameters of the correlation structure
cor_sph <- corSpher(value = c(15, 0.9), form = ~ range_num + row_num, nugget = T)

# Updating the model by plugin in the correlation function
m0_PGR_sph <- update(m0_bi_PGR, . ~ ., cor = cor_sph )

# Comparing non-spatial vs spatial model using LRT
anova(m0_bi_PGR, m0_PGR_sph) # sig
anova(m1_bi_PGR, m0_PGR_sph)

m1_PGR_sph <- update(m1_bi_PGR, . ~ ., cor = cor_sph )
anova(m1_bi_PGR, m1_PGR_sph) # not sig

anova(m0_bi_PGR, m1_bi_PGR) # sig

```

This output shows that there is a reduction of AIC (a gain on logLik) which is significant. This means that a model including the row and range is a better option to explain the variability of the data, but there is no need to include the correlation fucntion in the model.

 













######## check the bs_18 dataset############
```{r}
m0_bs_PGR <- lme(audpc ~ plot, bs_18.1, random = ~ 1 | rep, na.action = "na.omit")
anova(m0_bs_PGR, type = "marginal")
plot(m0_bs_PGR)
# there is trend left in the residuals
```


````{r}
# Getting residuals from the model
bs_18.1$resid_m0 <- residuals(m0_bs_PGR, type = "pearson")
bs_18.1_sp <- bs_18.1
coordinates(bs_18.1_sp) <- ~ range_num + row_num # give it coordiantes

# Spatial distribution
p_resid_m0_bs <- ggplot(bs_18.1, aes(x = range_num, y = row_num)) +
  geom_tile(aes(fill = resid_m0)) +
  theme_bw() + theme(panel.grid = element_blank()) +
  scale_fill_gradient(low = "darkblue", high = "orange") +
  ggtitle("Residual plot of 2018_BS population") + 
  labs(fill = "Residuals of model1") +
  xlab("row number") +
  ylab("range number")
p_resid_m0_bs

# pretty clear spatial trend in this plot
# there are more negative values at the small row number
```


```{r}
# Getting residuals from the model: no trend
vg0_bs<- Variogram(m0_bs_PGR, form = ~ range_num + row_num, resType = "pearson", maxDist = 50)
plot(vg0_bs, smooth = F, main="BS population variogram")
```

This variogram shows some structure, however it was not computed from a model including Xft and Yft as fixed predictors, thus we can not tell how much of that spatial structure can be attributed to the trend or it is structured random variation. To find this out we have to look at the variogram from a model including Xft and Yft as predictors.


```{r}
# Split-plot model with range and row as fixed predictors (for detrending)
m1_bs_PGR <- lme(audpc ~ plot, bs_18.1, random = ~ 1 | rep + 1|range_num + 1|row_num, na.action = "na.omit")

# Getting residuas from model 2
bs_18.1$resid_m1 <- residuals(m1_bs_PGR, type = "pearson")

# Getting variogram from the model
vg1_bs <- Variogram(m1_bs_PGR, form = ~ range_num + row_num, resType = "pearson", maxDist = 50)
plot(vg1_bs, smooth= F)
# the variogram seems to be much better
```

Now we can compare the effect of adding range and row for detrending purposes.
```{r}
p_resid_m1_bs <- ggplot(bs_18.1, aes(x = range_num, y = row_num)) +
  geom_tile(aes(fill = resid_m1)) +
  scale_fill_gradient(low = "darkblue", high = "orange") + 
  theme_bw() + theme(panel.grid = element_blank())
p_resid_m1_bs

grid.arrange( 
  p_resid_m0_bs + labs(title = "with trend"),
  p_resid_m1_bs + labs(title = "without trend"),
  nrow = 1,
   top="BS populationresidual plots"
)

# Comparing variograms
grid.arrange(
  plot(vg0_bs, smooth = F, main = "with trend"),
  plot(vg1_bs, smooth  =F, main = "without trend"),
  nrow = 1,
  top="BS semivariogram plots"
)
```


```{r}
# Defining intial parameters of the correlation structure
cor_sph_bs <- corSpher(value = c(25, 0.8), form = ~ range_num + row_num, nugget = T)

# Updating the model by plugin in the correlation function
m0_PGR_sph_bs <- update(m0_bs_PGR, . ~ ., cor = cor_sph_bs )
m1_PGR_sph_bs <- update(m1_bs_PGR, . ~ ., cor = cor_sph_bs )

# Comparing non-spatial vs spatial model using LRT
anova(m0_bs_PGR, m0_PGR_sph_bs) # not sig

anova(m1_bs_PGR, m1_PGR_sph_bs) # not sig

anova(m0_bi_PGR, m1_bi_PGR) # sig
```
based on the previous outputs of the test form differnt models, the model with row and range effect perform much better; but there is no need to include the correlation function after including row and range within the model.   



